Direct Spectral Detection: A Cheap Method to Detect and Characterize Binary Systems

### 2015-10-22 paper/ms.tex
Draft send to coauthors.

\label{sec:intro}
There has recently been a revival of interest in the multiplicity properties of intermediate-mass stars,  spurred largely by the detection of planets orbiting nearby $\sim 2 M_{\odot}$ stars on both wide \citep[e.g.][]{Lagrange2010, Marois2008} and close \citep{Johnson2011} orbits. In this context, binary companions are contaminants; companions complicate radial velocity planet searches because they necessitate simultaneous modeling of both stellar motions \citep[e.g.][]{Bergmann2015}. Likewise, companions complicate direct imaging planet searches by requiring either extremely high-contrast instrumentation \citep{Thalmann2014} or specialized coronagraphs \citep{Crepp2010}.

@@

 Radial velocity monitoring can easily find companions on very short-period orbits, but its sensitivity to low-mass companions drops as the physical separation increases. This is especially true for A- and B-type stars, where radial velocity precision is typically limited to $\sim 1 \rm km \  s^{-1}$ by their rotationally broadened lines. Worse, radial velocity monitoring techniques cannot characterize the companion unless the inclination is known or if the binary is double-lined. All three techniques have separation-dependent sensitivity, which introduces observational bias in any search for a parameter that changes with physical separation.

 One technique that is separation independent is to search directly for the composite spectrum of two stars. \citet{Burgasser2007} used single-epoch low-resolution spectroscopy to identify and characterize a brown dwarf binary system by fitting both spectra simultaneously. This method only works if the stars have a similar brightness but very different spectra, such that the spectral features from both components are visible and distinguishable. If, as in the case of binary systems with very large flux ratios, the companion spectrum is buried within the noise of the primary star then a different method is needed.

The direct spectral detection method (hereafter referred to as the DSD method), and variations thereof, has been used to search for binary companions to early B-stars \citep{Gullikson2013}, main sequence FGK-stars \citep{Kolbl2015}, young K-M stars \citep{Prato2002}, and even `Hot Jupiter' type planets \citep{Snellen2010, Brogi2012, deKok2013} orbiting FGK-stars. The method relies on the cross-correlation function (CCF) of a high-spectral-resolution spectrum of the primary star with a model spectrum for the expected companion. The CCF uses every pixel in the spectrum, and more importantly every spectral line in the secondary spectrum. A simple experiment with synthetic spectra containing increasing numbers of spectral lines (N) to noisy data shows that the CCF peak significance increases as $\sqrt{N}$. For high-resolution cross-dispersed \'echelle spectra, this amplification can reach several factors of 10, allowing the detection \emph{and characterization} of a secondary spectrum where the individual lines are completely buried in noise. Since the DSD method uses a seeing-limited spectrum of the primary star, its sensitivity is independent of separation inside $\sim 1 ''$ and can make use of small telescopes to detect high-contrast companions.

@@

\section{Direct Spectroscopic Detection}
\label{sec:method}

@@

Unlike most previous work, we focus on not only detecting but accurately characterizing the companion. We additionally optimize our technique for detecting cool companions to rapidly rotating early-type stars, for which it is very difficult to detect the reflex motion of the primary star. We fit and remove the telluric absorption using the TelFit code \citep{Gullikson2014}, and estimate an empirical primary star spectrum with a Gaussian smoothing filter applied to the telluric-corrected data. We chose to use a smoothing filter over subtracting model spectra for two reasons: the model spectra are a poor representation of the data, especially at the high signal-to-noise ratios that we use, and the smoothing filter removes any large-scale instrumental systematics that remain in the spectrum. The smoothing filter has a  window size ($w$) set by

@@

We estimate the expected companion temperature depending on whether it is part of a spectroscopic (Table \ref{tab:specdata}) or visual (Table \ref{tab:imagedata}) binary system. In the case of double-lined spectroscopic binaries, we use the ratio of the semi-amplitudes given in the 9th catalog of spectroscopic binary orbits \citep[SB9,][]{SB9} to estimate the mass ratio of the system. We convert the primary star spectral type from the Simbad database \citep{Simbad} to mass by interpolating Table 5 of \citet{Pecaut2013}. The mass ratio and primary mass gives an estimate of the companion mass, which we convert to temperature by interpolating the same table. Most of the directly-imaged binary systems do not have orbital information, so we use the magnitude difference published in the Washington Double Star catalog \citep[WDS, ][]{WDS}. We use the Simbad spectral type of the primary star and Table 5 of \citet{Pecaut2013} to estimate the primary star temperature ($T_1$) and radius ($R_1$). We then find the companion temperature that minimizes the following function for the companion temperature ($T_2$), given the observed magnitude difference ($\Delta m_{obs}$)
\begin{equation}
Q = (m(T_2, R_2) - m(T_1, R_1) - \Delta m_{obs})^2
\label{eqn:secteff}
\end{equation}
where $m(T, R)$ is the Vega magnitude of a star with temperature T and radius R. We use the pysynphot package\footnote{pysynphot is a python code package to perform synthetic photometry, and is available at this url: \url{https://pypi.python.org/pypi/pysynphot}} and a Kurucz model grid \citep{Castelli2003} to calculate $m(T, R)$, and assume the companion is on the main sequence to estimate its radius ($R_2$). For both spectroscopic and visual binary systems, we assume spectral type uncertainties of $\pm 1$ subtype on the primary stars, and propagate the uncertainties into uncertainty in the companion temperature. We include the binary system in the sample if the median companion temperature is $< 7500$ K.

We use the same set of instruments and settings for all observations throughout the three datasets. We use the CHIRON spectrograph \citep{CHIRON} on the 1.5m telescope at Cerro Tololo Inter-American Observatory for all southern targets. This spectrograph is an $R\equiv \lambda / \Delta \lambda = 80000$ cross-dispersed \'echelle spectrograph with wavelength coverage from 450 - 850 nm, and is fed by a $2.7''$ optical fiber. The data are automatically reduced with a standard CHIRON data reduction pipeline, but the pipeline leaves residuals of strong lines in adjacent orders. We bias-correct, flat-field and extract the spectra with the optimum extraction technique \citep{Horne1986} using standard IRAF\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.} tasks, and use the wavelength calibration from the pipeline reduced spectra.

@@

To measure the impact of both random and systematic noise on the parameter estimation, we created several hundred synthetic binary systems for each instrument used in our program. We made the synthetic binary systems by combining the early-type star spectra from Table \ref{tab:earlycal} with those of the late-type stars in Table \ref{tab:latecal} in every possible combination, provided both observations came from the same instrument. By combining actual observations of early-type and late-type stars, our synthetic binary observations retain any instrument-specific effects that may impact the temperature estimation. We scaled the flux of the late-type star such that the flux ratio ($F_{\rm secondary}/F_{\rm primary}$) is ten times larger than the expected flux ratio for main sequence components. The artificial brightening relative to a real binary system is to ensure that the temperature estimation uncertainties are separate from the overall sensitivity of the method (which we discuss in Section \ref{subsec:sensitivity}). We estimate the main sequence flux ratio from the published temperature of the late type star (given in Table \ref{tab:latecal}) and the published spectral types of the primaries available on Simbad \citep{Simbad}, and convert to temperature and luminosity by using Table 5 of \citet{Pecaut2013}.

@@

\begin{equation}
T_m = a + bT_a
\end{equation}
In Section \ref{sec:results} we invert this relation to determine better estimates for the actual companion temperature, given the measured temperature from Equation \ref{eqn:tmeas}.

@@

We compare our companion temperature measurements from the DSD method to the estimates described above in Figure \ref{fig:known}. There is overall excellent agreement between the temperatures, with most points falling within $1 \sigma$ of equality. We test for a bias ($\Delta$) between the measured temperatures ($T_m$) and expected temperatures ($T_a$) with the equations

\begin{eqnarray}
\Delta &= \sum_i(T_{m,i} - T_{a,i}) \\
\sigma_{\Delta}^2 &= \sum_i (\sigma_{T_{m,i}}^2 + \sigma_{T_{a,i}}^2)
\end{eqnarray}
which results in $\Delta = -580 \pm 890$ K. Our temperature measurements are consistent with the expected temperatures.

>>>>

\label{sec:intro}
There has recently been a revival of interest in the multiplicity properties of intermediate-mass stars,  spurred largely by the detection of planets orbiting nearby \mbox{$\sim2 M_{\odot}$} stars on both wide \citep[e.g.][]{Lagrange2010, Marois2008} and close \citep{Johnson2011} orbits. In this context, binary companions are contaminants; companions complicate radial velocity planet searches because they necessitate simultaneous modeling of both stellar motions \citep[e.g.][]{Bergmann2015}. Likewise, companions complicate direct imaging planet searches by requiring either extremely high-contrast instrumentation \citep{Thalmann2014} or specialized coronagraphs \citep{Crepp2010}.

@@

 Radial velocity monitoring can easily find companions on very short-period orbits, but its sensitivity to low-mass companions drops as the physical separation increases. This is especially true for A- and B-type stars, where radial velocity precision is typically limited to $\sim 1 \rm km \  s^{-1}$ by their rotationally broadened lines. Worse, radial velocity monitoring techniques cannot characterize the companion unless the inclination is known or if the companion spectral lines are also visible. All three techniques have separation-dependent sensitivity, which introduces observational bias in any search for a parameter that changes with physical separation.

 One technique that is separation independent is to search directly for the composite spectrum of two stars. \citet{Burgasser2007} used single-epoch low-resolution spectroscopy to identify and characterize a brown dwarf binary system by fitting both spectra simultaneously. This method only works if the stars have a similar brightness but very different spectra, such that the spectral features from both components are visible and distinguishable. If, as in the case of binary systems with very large flux ratios, the companion spectrum is buried within the noise of the primary star, a different method is needed.

The direct spectral detection method (hereafter referred to as the DSD method), and variations thereof, has been used to search for binary companions to early B-stars \citep{Gullikson2013}, main sequence FGK-stars \citep{Kolbl2015}, young K-M stars \citep{Prato2002}, and even `Hot Jupiter' type planets \citep{Snellen2010, Brogi2012, deKok2013} orbiting FGK-stars. The method relies on the cross-correlation function (CCF) of a high-spectral-resolution spectrum of the primary star with a model spectrum for the expected companion. The CCF uses every pixel in the spectrum, and more importantly every spectral line in the secondary spectrum. A simple experiment with synthetic spectra containing increasing numbers of spectral lines (N) in noisy data shows that the CCF peak significance increases as $\sim\sqrt{N}$. For high-resolution cross-dispersed \'echelle spectra, this amplification can reach several factors of 10, allowing the detection \emph{and characterization} of a secondary spectrum where the individual lines are completely buried in noise. Since the DSD method uses a seeing-limited spectrum of the primary star, its sensitivity is independent of separation inside $\sim 1 ''$ and can make use of small telescopes to detect high-contrast companions.

@@

\section{Direct Spectral Detection Method}
\label{sec:method}

@@

Unlike most previous work, we focus on not only detecting but accurately characterizing the companion. We additionally optimize our technique for detecting cool companions to rapidly rotating early-type stars, for which it is very difficult to detect the reflex motion of the primary star. We fit and remove the telluric absorption using the TelFit code \citep{Gullikson2014}, and estimate an empirical primary star spectrum with a Gaussian smoothing filter applied to the telluric-corrected data. We chose to use a smoothing filter over subtracting model spectra for two reasons: first, the model spectra are a poor representation of the data, especially at the high signal-to-noise ratios that we use, and so leave very large-scale features in the residual spectrum. Second, the smoothing filter will also remove any large-scale instrumental systematics in the spectrum. We use a smoothing filter with a  window size ($w$) set by

@@

We estimate the expected companion temperature depending on whether it is part of a spectroscopic (Table \ref{tab:specdata}) or visual (Table \ref{tab:imagedata}) binary system. In the case of double-lined spectroscopic binaries, we use the ratio of the semi-amplitudes given in the 9th catalog of spectroscopic binary orbits \citep[SB9,][]{SB9} to estimate the mass ratio of the system. We convert the primary star spectral type from the Simbad database \citep{Simbad} to mass by interpolating Table 5 of \citet{Pecaut2013}. The mass ratio and primary mass gives an estimate of the companion mass, which we convert to temperature by interpolating the same table. Most of the directly-imaged binary systems do not have orbital information, so we use the magnitude difference published in the Washington Double Star catalog \citep[WDS, ][]{WDS}. We use the Simbad spectral type of the primary star and Table 5 of \citet{Pecaut2013} to estimate the primary star temperature ($T_1$) and radius ($R_1$). We then find the companion temperature that minimizes the following function for the companion temperature ($T_2$), given the observed magnitude difference ($\Delta m_{obs}$)
\begin{equation}
Q = (m(T_2, R_2) - m(T_1, R_1) - \Delta m_{obs})^2
\label{eqn:secteff}
\end{equation}
where $m(T, R)$ is the Vega magnitude of a star with temperature T and radius R. We use the pysynphot package\footnote{pysynphot is a python code package to perform synthetic photometry, and is available at this url: \url{https://pypi.python.org/pypi/pysynphot}} and a Kurucz model grid \citep{Castelli2003} to calculate $m(T, R)$, and assume the companion is on the main sequence to estimate its radius ($R_2$). For both spectroscopic and visual binary systems, we assume spectral type uncertainties of $\pm 1$ subtype on the primary stars, and propagate the uncertainties into uncertainty in the companion temperature. We include the binary system in the sample if the expected companion temperature is $< 7500$ K.

We use the same set of instruments and settings for all observations throughout the three datasets. We use the CHIRON spectrograph \citep{CHIRON} on the 1.5m telescope at Cerro Tololo Inter-American Observatory for most southern targets. This spectrograph is an $R\equiv \lambda / \Delta \lambda = 80000$ cross-dispersed \'echelle spectrograph with wavelength coverage from 450 - 850 nm, and is fed by a $2.7''$ optical fiber. The data are automatically reduced with a standard CHIRON data reduction pipeline, but the pipeline leaves residuals of strong lines in adjacent orders. We therefore bias-correct, flat-field and extract the spectra with the optimum extraction technique \citep{Horne1986} using standard IRAF\footnote{IRAF is distributed by the National Optical Astronomy Observatories, which are operated by the Association of Universities for Research in Astronomy, Inc., under cooperative agreement with the National Science Foundation.} tasks, and use the wavelength calibration from the pipeline reduced spectra.

@@

To measure the impact of both random and systematic noise on the parameter estimation, we created several hundred synthetic binary systems for each instrument used in our program. We made the synthetic binary systems by combining the early-type star spectra from Table \ref{tab:earlycal} with those of the late-type stars in Table \ref{tab:latecal} in every possible combination, provided both observations came from the same instrument. By combining actual observations of early-type and late-type stars, our synthetic binary observations retain any instrument-specific effects that may impact the temperature estimation. We scaled the flux of the late-type star such that the flux ratio ($F_{\rm secondary}/F_{\rm primary}$) is ten times larger than the expected flux ratio for main sequence components. The artificial brightening relative to a real binary system is to ensure that the temperature estimation uncertainties are separate from the overall sensitivity of the method, which we discuss in Section \ref{subsec:sensitivity}. We estimate the main sequence flux ratio from the published temperature of the late type star (given in Table \ref{tab:latecal}) and the published spectral types of the primaries available on Simbad \citep{Simbad}, and convert to temperature and luminosity by using Table 5 of \citet{Pecaut2013}.

@@

\begin{equation}
T_m = a + bT_a
\end{equation}
In Section \ref{sec:results} we invert this relation to determine the actual companion temperature, given the measured temperature from Equation \ref{eqn:tmeas}.

@@

We compare our companion temperature measurements from the DSD method to the estimates described above in Figure \ref{fig:known}. There is overall excellent agreement between the temperatures, with 5/6 falling within $1 \sigma$ of equality. We test for a bias ($\Delta$) between the measured temperatures ($T_m$) and expected temperatures ($T_a$) with the equations

\begin{eqnarray}
\Delta &= \sum_i(T_{m,i} - T_{a,i}) \\
\sigma_{\Delta}^2 &= \sum_i (\sigma_{T_{m,i}}^2 + \sigma_{T_{a,i}}^2)
\end{eqnarray}
which results in $\Delta = -580 \pm 770$ K. Our temperature measurements are consistent with the expected temperatures.
